### 2018-01-25 submission/manuscript.Rmd
ada edits on abstract

**Background.** An increasing body of literature suggests that there is a crucial role for the microbiota in colorectal cancer (CRC) pathogenesis. Important drivers within this context have ranged from individual microbes to the whole community. Our study expands on a recent meta-analysis investigating microbial biomarkers for CRC by testing the hypothesis that the bacterial community has important associations to both early (adenoma) and late (carcinoma) stage disease. To test this hypothesis we examined both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) across 14 different studies.

**Results.** Fecal samples had a significant decrease from control to adenoma to carcinoma for both Shannon diversity and evenness after correcting for study effect and variable region sequenced (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). This reduction in evenness resulted in small increases in relative risk for adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only resulted in an increased relative risk for carcinoma (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`). Mouth-associated genera were consistently in the top 5 OTUs or genera with the highest relative risk of adenoma and carcinoma for both stool and tissue. Using the whole community versus selected generea with the highest and lowest relative risk, to build a prediction model, resulted in similar classification success, based on Area Under the Curve (AUC), for both adenoma and carcinoma using fecal and tissue samples. The most important groups within the full community models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies. Overall, there were less associations between the microbiota and adenoma and one reason why this may be is that most studies were only adequately powered for large effect sizes.

**Conclusions.** This data provides support for the importance of the bacterial community to both adenoma and carcinoma genesis. The evidence collected within this study on the role of the microbiota in CRC pathogenesis shows stronger associations between carcinoma then adenoma and this may be in part due to the low power to detect more subtle changes in the majority of studies that have been performed to date.

>>>>

**Background.** An increasing body of literature suggests that there is a crucial role for the microbiota in colorectal cancer (CRC) pathogenesis. Important drivers within this context have ranged from individual microbes to the whole community. Our study expands on a recent meta-analysis investigating microbial biomarkers for CRC by testing the hypothesis that the bacterial community has important associations to both early (adenoma) and late (carcinoma) stage disease. To test this hypothesis we examined both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) across 14 previously published 16S rRNA gene sequencing studies on CRC and the microbiota.

**Results.** Fecal samples had a significant decrease from those without disease to adenoma (less severe disease) to carcinoma (severe disease) for both Shannon diversity and evenness after correcting for study effect and variable region sequenced (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). This reduction in evenness translated into small increases in relative risk for adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma stages of CRC (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only translated into an increased relative risk for developing carcinomas (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`). Increases in mouth-associated microbes were commonly in the top 5 most significantly increased relative risk of adenoma and carcinoma for both stool and tissue samples. A prediction model for adenoma and carcinoma was built using either the whole community or selected genera with highest and lowest relative risk from fecal and tissue samples. Both approaches resulted in similar classification success according to Area Under the Curve (AUC) regardless of whether genera or OTUs were used to build the model. The most important groups within the full community models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies. Although a number of associations between the microbiota and CRC were identified, the majority of studies that we used in this meta-analysis were only individually adequately powered for large effect sizes.

**Conclusions.** These data provide support for the importance of the bacterial community to both adenoma and carcinoma genesis. The evidence collected within this study on the role of the microbiota in CRC identifies a number of correlations that may not have been detected because of the low power associated with the majority of studies that have been performed to date.

### 2018-01-25 submission/manuscript.tex
ada edits on abstract

\textbf{Background.} An increasing body of literature suggests that
there is a crucial role for the microbiota in colorectal cancer (CRC)
pathogenesis. Important drivers within this context have ranged from
individual microbes to the whole community. Our study expands on a
recent meta-analysis investigating microbial biomarkers for CRC by
testing the hypothesis that the bacterial community has important
associations to both early (adenoma) and late (carcinoma) stage disease.
To test this hypothesis we examined both feces (n = 1737) and tissue
(492 total samples from 350 individuals) across 14 different studies.

\textbf{Results.} Fecal samples had a significant decrease from control
to adenoma to carcinoma for both Shannon diversity and evenness after
correcting for study effect and variable region sequenced (P-value
\textless{} 0.05). This reduction in evenness resulted in small
increases in relative risk for adenoma (P-value = 0.032) and carcinoma
(P-value = 0.00034) while the reduction in Shannon diversity only
resulted in an increased relative risk for carcinoma (P-value = 0.0047).
Mouth-associated genera were consistently in the top 5 OTUs or genera
with the highest relative risk of adenoma and carcinoma for both stool
and tissue. Using the whole community versus selected generea with the
highest and lowest relative risk, to build a prediction model, resulted
in similar classification success, based on Area Under the Curve (AUC),
for both adenoma and carcinoma using fecal and tissue samples. The most
important groups within the full community models consistently belonged
to genera such as \emph{Ruminococcus}, \emph{Bacteroides}, and
\emph{Roseburia} across studies. Overall, there were less associations
between the microbiota and adenoma and one reason why this may be is
that most studies were only adequately powered for large effect sizes.

\textbf{Conclusions.} This data provides support for the importance of
the bacterial community to both adenoma and carcinoma genesis. The
evidence collected within this study on the role of the microbiota in
CRC pathogenesis shows stronger associations between carcinoma then
adenoma and this may be in part due to the low power to detect more
subtle changes in the majority of studies that have been performed to
date.

>>>>

\textbf{Background.} An increasing body of literature suggests that
there is a crucial role for the microbiota in colorectal cancer (CRC)
pathogenesis. Important drivers within this context have ranged from
individual microbes to the whole community. Our study expands on a
recent meta-analysis investigating microbial biomarkers for CRC by
testing the hypothesis that the bacterial community has important
associations to both early (adenoma) and late (carcinoma) stage disease.
To test this hypothesis we examined both feces (n = 1737) and colon
tissue (492 total samples from 350 individuals) across 14 previously
published 16S rRNA gene sequencing studies on CRC and the microbiota.

\textbf{Results.} Fecal samples had a significant decrease from those
without disease to adenoma (less severe disease) to carcinoma (severe
disease) for both Shannon diversity and evenness after correcting for
study effect and variable region sequenced (P-value \textless{} 0.05).
This reduction in evenness translated into small increases in relative
risk for adenoma (P-value = 0.032) and carcinoma stages of CRC (P-value
= 0.00034) while the reduction in Shannon diversity only translated into
an increased relative risk for developing carcinomas (P-value = 0.0047).
Increases in mouth-associated microbes were commonly in the top 5 most
significantly increased relative risk of adenoma and carcinoma for both
stool and tissue samples. A prediction model for adenoma and carcinoma
was built using either the whole community or selected genera with
highest and lowest relative risk from fecal and tissue samples. Both
approaches resulted in similar classification success according to Area
Under the Curve (AUC) regardless of whether genera or OTUs were used to
build the model. The most important groups within the full community
models consistently belonged to genera such as \emph{Ruminococcus},
\emph{Bacteroides}, and \emph{Roseburia} across studies. Although a
number of associations between the microbiota and CRC were identified,
the majority of studies that we used in this meta-analysis were only
individually adequately powered for large effect sizes.

\textbf{Conclusions.} These data provide support for the importance of
the bacterial community to both adenoma and carcinoma genesis. The
evidence collected within this study on the role of the microbiota in
CRC identifies a number of correlations that may not have been detected
because of the low power associated with the majority of studies that
have been performed to date.
